---
title: "P8105 Homework 3"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(knitr)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


# Problem 2

## Loading, tidying, merging, and organizing the data sets
```{r}
accel_df = read_csv(file = "./hw3_data/nhanes_accel.csv") |> 
  janitor::clean_names()

covar_df = read_csv(file = "./hw3_data/nhanes_covar.csv", skip = 4) |> 
  janitor::clean_names()
```


```{r}
nhanes_df = accel_df |> 
  left_join(covar_df, by = "seqn") |> 
  filter(age >= 21) |>
  mutate(
    sex = as.character(sex), 
    sex = case_match(
      sex,
      "1" ~ "male",
      "2" ~ "female"),
    education = as.character(education),
    education = case_match(
      education,
      "1" ~ "Less than high school",
      "2" ~ "High school equivalent",
      "3" ~ "More than high school"
    )
  ) |>
  distinct()
```


## Number of men and women in each education category
```{r}
nhanes_df |> 
  group_by(education, sex) |> 
  summarize(
    count = n(), .groups = 'drop') |> 
  pivot_wider(
    names_from = education, 
    values_from = count) |> 
  knitr::kable(
    caption = "Number of Men and Women in Each Education Category"
  )
```

According to the table, participants in the 'More than high school' education category make up most of the total population compared to the 'High school equivalent' and 'Less than high school' education categories. Within the 'High school equivalent' category, there are 13 more males than females, 1 more female than males within the 'Less than high school' category, and 3 more females than males within the 'More than high school' category. 

### Visualization
```{r}
nhanes_df |> 
  group_by(sex, education, age) |> 
  ggplot(aes(x = education, y = age, color = sex)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Age by Sex and Education",
    x = "Education",
    y = "Age"
  )
```

A boxplot is used to visualize the median age between men and women across the different education categories as it is much easier to interpret for readers than other plots. The plot shows that females have a higher median age in the 'High school equivalent' category and the 'Less than high school' category. However, men have a slightly higher median age in the 'More than high school category.

## Total activities vs. age
```{r}
total_df = 
  nhanes_df |> 
  mutate(
    total_activity = rowSums(across(c(min1:min1440)))
  ) |> 
  select(seqn:education, total_activity)
```

```{r}
total_df |> 
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_grid(~ education) +
  labs(
    title = "Total Activity vs. Age by Gender and Education Level",
       x = "Age (years)",
       y = "Total Activity over 24 hours")
```


## 24-hour activity time courses for each education level
```{r}
panel_df = 
  nhanes_df |> 
  select(-seqn, -bmi, -age) |> 
  group_by(sex, education) |> 
  summarize(across(min1:min1440, mean)) |> 
  pivot_longer(
    min1:min1440, 
    names_to = "minutes",
    values_to = "activity"
  )
```

```{r}
panel_df |> 
  ggplot(aes(x = minutes, y = activity, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_grid(education ~ .) +
  labs(
    title = "Progression of activity over 24 hours"
  )
```



# Problem 3

## Importing, cleaning, and tidying data on rides taken on the NYC Citi Bike system
```{r}
jan_2020 = read_csv(file = "./hw3_data/Jan_2020_Citi.csv") |> 
  drop_na() |> 
  janitor::clean_names() |> 
  mutate(month_year = "jan_20") |> 
  select(month_year, everything())

jan_2024 = read_csv(file = "./hw3_data/Jan_2024_Citi.csv") |> 
  drop_na() |> 
  janitor::clean_names() |> 
  mutate(month_year = "jan_24") |> 
  select(month_year, everything())

july_2020 = read_csv(file = "./hw3_data/July_2020_Citi.csv") |>  
  drop_na() |> 
  janitor::clean_names() |> 
  mutate(month_year = "jul_20") |> 
  select(month_year, everything())

july_2024 = read_csv(file = "./hw3_data/July_2024_Citi.csv") |>
  drop_na() |> 
  janitor::clean_names() |> 
  mutate(month_year = "jul_24") |> 
  select(month_year, everything())

citibike_df = 
  bind_rows(jan_2020, jan_2024, july_2020, july_2024)
```


## Total number of rides in each combination of year and month, separating casual riders and Citi Bike members
```{r}
citibike_df |> 
  group_by(month_year, member_casual) |> 
  summarize(
    count = n(), .groups = 'drop') |> 
  pivot_wider(
    names_from = "member_casual", 
    values_from = "count") |> 
  knitr::kable(
    caption = "Total number of rides by casual riders and Citi Bike members"
  )
```


## 5 most popular starting stations for July 2024
```{r}
july_2024 |> 
  group_by(start_station_name) |> 
  count(start_station_name, name = "ride_count") |> 
  arrange(desc(ride_count)) |> 
  filter(ride_count >= 145) |> 
  knitr::kable(
    col.names = c("Origin Station Name", "Number of Rides")
  )
```


## Investigating the effects of day of the week, month, and year on median ride duration
```{r}
citibike_df |> 
  group_by(month_year, weekdays) |> 
  summarize(
    median = median(duration, na.rm = TRUE)) |> 
  ggplot(aes(x = weekdays, y = median, color = month_year)) +
  geom_point() +
  facet_grid(. ~ month_year) +
  labs(
    title = "Ride duration over week and 4 timepoints",
    x = "Days of the Week", 
    y = "Median of Ride Duration"
  )
```


## Impact of month, membership status, and bike type on the ditribution of ride duration
```{r}
citibike_df |> 
  group_by(month_year, rideable_type, member_casual) |> 
  select(month_year, rideable_type, member_casual, duration) |> 
  filter(month_year %in% c("jan_24", "jul_24")) |> 
  ggplot(aes(x = month_year, y = duration, fill = rideable_type)) +
  geom_violin() +
  facet_grid(member_casual ~ month_year) +
  labs(
    title = "Electric bikes ridership based on types of riders and time",
    x = "Month",
    y = "Ride Duration",
    color = "Types of Riders"
  )
```

